A pipeline to analyse time-course gene expression data

### 2020-07-15 scripts/manuscript.Rmd
continuing to transition manuscript

@@

genes_to_keep = row.names(
    log_fold_change_max[
	(rowSums(log_fold_change_max > 2) > 0) &
	(fishers_qval < 0.05), ])
# Keep the data corresponding to the genes of interest in another variable.
# by subsetting the `moanin_model`, which contains the data.
y = moanin_model[genes_to_keep,]
```

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(y, n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

```{r fig.height=6, fig.width=6}
plot_splines_data(moanin_model,
    data=kmeans_clusters$centroids, 
    colors=ann_colors$Group,
    smooth=TRUE)
```

@@

```{r genesVsCentroids, fig.height=6, fig.width=6}
cluster_of_interest = 2
cluster2Genes = names(
    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
plot_splines_data(moanin_model, data=dataToPlot, 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

@@

```{r visualizeScores, fig.height=6, fix.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
# Get the best score and best label for all of the genes
# This time without filtering labels
# We can give the previous calculated scores to `previous_scores` to save time
unfiltered_scores<-splines_kmeans_score_and_label(moanin_model,kmeans_clusters,proportion_genes_to_label=1, previous_scores=scores_and_labels$scores)
best_score = rowMin(scores_and_labels$scores)
best_label = unfiltered_scores$labels
par(mfrow=c(3, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(best_score[best_label==cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
    abline(v=scores_and_labels$score_cutoff, col="red", lwd=3, lty=2)
}

@@

```{r topScoreInCluster2, fig.height=6, fig.width=6, fig.cap=paste("Top scoring genes in Cluster",cluster_of_interest)}
# order them by their score in cluster
ord = order(scores_and_labels$scores[genes_in_cluster2,
                     cluster_of_interest])
cluster2Score = genes_in_cluster2[ord]
dataToPlot = rbind(data[cluster2Score[1:4], ],
    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
plot_splines_data(moanin_model, data=dataToPlot, 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

res = aheatmap(
    assay(moanin_model[genes_to_plot,ord]),
    Colv=NA,
    color="YlGnBu",
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=FALSE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(raw)"),
    treeheight=0, legend=FALSE)

# Now use the results of the previous call to aheatmap to reorder the genes.
aheatmap(
    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
    Colv=NA,
    Rowv=NA,
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=TRUE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
    treeheight=0)

>>>>

The `Moanin` class extends the `SummarizedExperiment` class of Bioconductor, so that the data as well as the meta data and our spline information are all held in one object. This means that the object can be subsetted just like the data matrix, and the corresponding meta data and basis function evaluations will be similarly subsetted:

```{r demonstrateSubsetting}
dim(moanin_model)
moanin_model[1:10,1:3]
```

@@

genes_to_keep = row.names(
    log_fold_change_max[
	(rowSums(log_fold_change_max > 2) > 0) &
	(fishers_qval < 0.05), ])
# Keep the data corresponding to the genes of interest in another variable.
# by subsetting the `moanin_model`, which contains the data.
de_moanin_model = moanin_model[genes_to_keep,]
```

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(de_moanin_model, n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

```{r fig.height=6, fig.width=6}
plot_splines_data(de_moanin_model,
    data=kmeans_clusters$centroids, 
    colors=ann_colors$Group,
    smooth=TRUE)
```

@@

```{r genesVsCentroids, fig.height=6, fig.width=6}
cluster_of_interest = 2
cluster2Genes = names(
    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
dataToPlot=rbind(assay(de_moanin_model[cluster2Genes[3:6],]), 
    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
plot_splines_data(de_moanin_model,  data=dataToPlot, 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

We can also overlay the centroid on the data, after rescaling the data to be between 0-1, like the centroids

```{r}
de_moanin_rescale = de_moanin_model
assay(de_moanin_rescale) = rescale_values(data=assay(de_moanin_model))
plot_splines_data(de_moanin_rescale,  data=kmeans_clusters$centroids[cluster_of_interest,], 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    data_smooth_only=TRUE, subset_data=cluster2Genes[3:6],
    mar=c(1.5,2.5,2,0.1))
```
**EAP** I added the functionality to plot centroid over data, rather than just fit to data, but I'm not sure it makes much sense to include this. The `rescale_values` function is too sensitive to outliers. Would be better to rescale each centroid to scale of gene. If we had a helper function to do that, could easily add that to plot_splines_data (made issue on moanin), but probably right now should remove this.

@@

```{r visualizeScores, fig.height=6, fix.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
# Get the best score and best label for all of the genes
# This time without filtering labels
# We can give the previous calculated scores to `previous_scores` to save time
unfiltered_scores<-splines_kmeans_score_and_label(moanin_model,kmeans_clusters,proportion_genes_to_label=1, previous_scores=scores_and_labels$scores)
best_score = rowMin(scores_and_labels$scores)
best_label = unfiltered_scores$labels

par(mfrow=c(3, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(best_score[best_label==cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
    abline(v=scores_and_labels$score_cutoff, col="red", lwd=3, lty=2)
}

@@

```{r topScoreInCluster2, fig.height=6, fig.width=6, fig.cap=paste("Top scoring genes in Cluster",cluster_of_interest)}
# order them by their score in cluster
ord = order(scores_and_labels$scores[genes_in_cluster2,
                     cluster_of_interest])
cluster2Score = genes_in_cluster2[ord]
dataToPlot = rbind([cluster2Score[1:4], ],
    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
plot_splines_data(moanin_model, data=dataToPlot, 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

res = aheatmap(
    assay(de_moanin_model[genes_to_plot,ord]),
    Colv=NA,
    color="YlGnBu",
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=FALSE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(raw)"),
    treeheight=0, legend=FALSE)

# Now use the results of the previous call to aheatmap to reorder the genes.
aheatmap(
    rescale_values(de_moanin_model[genes_to_plot,ord])[res$rowInd,],
    Colv=NA,
    Rowv=NA,
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=TRUE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
    treeheight=0)
